---
title: "Fisher wave"
author: "Clara Rehmann"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(data.table)
library(scales)
```

## Fisher wave

If we let an allele spread across a landscape at a given velocity (a la Fisher's wave of advance of advantageous genes), what does the relationship between frequency and area look like?

```{r wave}
allele_area <- function(r, # radius of circle
                        landscape_width = 25
                        ) {
  a=landscape_width/2 # maximum radius before allele exits circle
  if (r <= a) { # circle is completely within the landscape
    area=pi*(r**2)
  }
  else if (r >= (sqrt(2*(a**2)))) { # circle is completely outside the landscape
    area=(a*2)**2
  }
  else { # circle exits the landscape - do some fancy trig
         # https://math.stackexchange.com/questions/1450961/overlapping-area-between-a-circle-and-a-square
    theta=acos(a/r)
    area=(((pi-(4*theta))*(r**2))+(4*a*r*sin(theta)))
  }
  return(area)
}

allele_frequency <- function(a, #area of allele
                             f = 1, #frequency of allele inside area
                             landscape_width = 25
                            ){
  A=landscape_width**2 # size of landscape
  return(f*(a/A))
}

Fisher_wave <- function(M, # migration rate
                        S, # selection coefficient
                        landscape_width = 25) {
  landscape_area=landscape_width**2
  rmax=landscape_width/2 # radius @ which circle starts escaping the landscape
  Rmax=sqrt(2*(rmax**2)) # radius @ which circle encompasses landscape
  V=2*sqrt(S*M) # allele velocity
  radii=seq(0, Rmax, V) # radius of the allele area circle after each generation
  
  areas <- sapply(radii, allele_area)
  frequencies <- sapply(areas, allele_frequency)
  return(data.table(areas, frequencies))
}
```

We're making two assumptions at first:
First, we'll assume an infinite population across the landscape so that we can approximate allele frequency as the proportion of landscape area occupied by the allele...
Second, we're approximating the wave *without* heterozygotes - so everywhere the radius of the circle is, the allele is fixed, and everywhere it isn't, it's absent.

```{r wave-test}
M=0.01
S=0.1 # selection coefficient
ggplot(Fisher_wave(M, S)) + geom_line(aes(x=frequencies, y=areas))
```

Shit, that is just linear. What's going on here?

(A) The curve of the frequency/area joint distribution from our sims depends on the fact that population density increases where the allele is more frequent

(B) The heterozygotes matter

Option B is the more straightforward one to check out, so let's modify the code to include a second circle bound for the heterozygotes at the front of the wave.
Within this bound (but outside of the bound of fixation), we'll approximate the frequency to be 0.5.

```{r}
M = 0.1
S = 0.05
Fisher_wave_H <- function(M, # migration rate
                          S, # selection coefficient
                          landscape_width = 25) {
  landscape_area=landscape_width**2
  rmax=landscape_width/2 # radius @ which circle starts escaping the landscape
  Rmax=sqrt(2*(rmax**2)) # radius @ which circle encompasses landscape
  
  lambd=sqrt(M/S)
  V=M*sqrt(S) # allele velocity
  #W =1/sqrt(S*M)
  #W=4*lambd
  #W = sqrt((2*M)/S)
  W=1/(4*sqrt(S)) # width of wavefront?????????????????????????????????????????????????
  hom_radii=seq(0, Rmax, V) # radius of the homozygous allele area circle after each generation
  het_radii=seq(0, Rmax+W, V) # radius of the heterozygous allele area circle after each generation
  hom_radii <- c(rep(0, length(het_radii) - length(hom_radii)), hom_radii) # add het only generations
  
  
  hom_areas <- sapply(hom_radii, allele_area)
  het_areas <- sapply(het_radii, allele_area)
  allele_areas <- het_areas
  het_areas <- het_areas - hom_areas
  frequencies <- sapply(hom_areas, allele_frequency) + 
    (sapply(het_areas, allele_frequency, 0.5))
  #frequencies <- rescale(frequencies)
  return(data.table(allele_areas, frequencies))
}

ggplot(Fisher_wave_H(M, S)) + geom_line(aes(x=frequencies, y=allele_areas))
```

YEAAAAAAAAAAAAA

does it change with selection?

```{r}
for (s in c(0.0001, 0.001, 0.01, 0.1)){
  df <- Fisher_wave_H(0.1, s)
  df$s <- s
  df$tick <- seq(nrow(df))
  assign(paste0('S_',s), df)
}

df <- rbind(`S_1e-04`, S_0.001, S_0.01, S_0.1)
ggplot(df) + theme_bw() + geom_line(aes(x=frequencies, y=allele_areas, group=s, color=s)) + scale_color_continuous(trans='log10')
```

```{r}
for (s in c(0.0001, 0.001, 0.01, 0.1)){
  df <- Fisher_wave_H(1, s)
  df$s <- s
  assign(paste0('S_',s), df)
}

df <- rbind(`S_1e-04`, S_0.001, S_0.01, S_0.1)
ggplot(df) + theme_bw() + geom_line(aes(x=frequencies, y=allele_areas, group=s, color=s)) + scale_color_continuous(trans='log10')
```


```{r}
for (s in c(0.0001, 0.001, 0.01, 0.1)){
  df <- Fisher_wave_H(0.1, s)
  df$s <- s
  assign(paste0('S_',s), df)
}

df <- rbind(`S_1e-04`, S_0.001, S_0.01, S_0.1)
ggplot(df) + theme_bw() + geom_line(aes(x=frequencies, y=allele_areas, group=s, color=s)) + scale_color_continuous(trans='log10')
```


```{r}
plot(df[df$s==0.0001]$frequencies)

```

So it's close, but the 0.5 approximation is still messing stuff up. Let's try approximating the frequency in that area as the volume of a cone...


```{r fancy-het}
allele_hom_frequency <- function(a, #area of allele
                                 f = 1, #frequency of allele inside area
                                 landscape_width = 25
                                 ){
  A=landscape_width**2 # size of landscape
  return(f*(a/A))
}

allele_het_frequency <- function(hetR, # radius of heterozygotes
                                 homR, # radius of homozygotes
                                 W, # width of wave
                                 landscape_width=25){
    if (hetR == 0){ return(0) }
    rmax = hetR # maximum value of het wave
    a=landscape_width/2 # maximum radius before allele exits circle
    # function for hypotenuse
    slope=-1/(rmax-homR) # slope of hypotenuse
    intercept=rmax/(rmax+homR) # intercept of hypotenuse
    f <- function(x){
      return(slope*x+intercept)
    }
    
    Istart <- homR # start of interval
    Istop <- rmax # end of interval
    allelevolume <- pi * integrate(function(x) (f(x))^2, Istart, Istop)$value

    if (rmax > a){ # circle exceeds landscape
      extravol <- pi * integrate(function(x) (f(x))^2, a, Istop)$value
      allelevolume <- allelevolume - extravol
    }
    
    # full volume of cylinder ring
    c <- function(x){
      return(2 * pi * x)
    }
    cyvolume <- integrate(c, homR, rmax)$value
    if (rmax > a){ # circle exceeds landscape
      extravol <- integrate(c, a, Istop)$value
      cyvolume <- cyvolume - extravol
    }
    area = allelevolume/cyvolume 
    return(area)
}

allele_cone <- function(het_radius, # radius of heterozygotes
                        hom_radius, # radius of homozygotes
                        hom_area, # area of homozygotes
                landscape_width = 25){
  if (het_radius == 0){ return(0) }
  rmax = landscape_width/2 # radius the cone starts exiting the circle
  Rmax=sqrt(2*(rmax**2)) # radius @ which circle encompasses landscape
  slope=-1/(het_radius-hom_radius) # slope of hypotenuse
  intercept=-slope * het_radius # intercept of hypotenuse
  fx <- function(x){
    return(x*slope + intercept)
  }
  fy <- function(y){
    return((y - intercept)/slope)
  }
  if (het_radius <= rmax){
  # volume of het cone
  allelevolume <- pi * het_radius^2 * (intercept / 3)
  # cut off top of cone
  alleletop <- pi * hom_radius^2 * ((intercept-1)/3)
  allelevolume <- allelevolume-alleletop
  }
  if ((het_radius > rmax) & (het_radius <= Rmax)) {
    allelevolume <- pi * het_radius^2 * (intercept / 3)
    # cut off top of cone
    alleletop <- pi * hom_radius^2 * ((intercept-1)/3)
    allelevolume <- allelevolume-alleletop
    h = intercept # height of whole cone
    r = het_radius # radius of whole cone
    d = rmax # distance from origin to slice
    # cone slices https://math.stackexchange.com/questions/1660071/how-to-calculate-volume-of-a-right-circular-cones-hyperbola-segment
    slicevol <- (h/3) * 
                ((-2 * d * sqrt(r^2 - d^2)) +
                      ((d^3 / r) * log(sqrt(r^2 - d^2) + r)) +
                      ((r^2 * acos(d/r)) - ((d^3 / r) * log(d))))
    h1 <- intercept - 1
    r1 <- fy(1)
    if (h1 > 0){
    #cone volume above y=1 was already subtracted, so remove that from the slice
      outside <- h1/3 * 
                 ((-2 * d * sqrt(r1^2 - d^2)) +
                      ((d^3 / r1) * log(sqrt(r1^2 - d^2) + r1)) +
                      (r1^2 * acos(d/r1) - ((d^3 / r1) * log(d))))
      if (is.na(outside)){slicevol = slicevol}
      else {slicevol <- slicevol - outside}
    }
    allelevolume <- allelevolume - 4*slicevol
  }
    if (het_radius > Rmax){
      # full volume of 25x25 box beneath cone
      circleintersectY <- fx(Rmax)
      boxvol <- 25 * 25 * circleintersectY
      # cone on top
      coneheight <- intercept - circleintersectY
      conevol <- pi * Rmax^2 * (coneheight/3)
      conetop <- pi * hom_radius^2 * ((intercept-1)/3)
      conevol <- conevol - conetop
      # slices from the outside
      d <- rmax
      r <- Rmax
      h <- coneheight
      slicevol <- (h/3) * 
                ((-2 * d * sqrt(r^2 - d^2)) +
                      ((d^3 / r) * log(sqrt(r^2 - d^2) + r)) +
                      ((r^2 * acos(d/r)) - ((d^3 / r) * log(d))))
      h1 <- intercept - 1
      r1 <- fy(1)
      if (h1 > 0){
      # cone volume above y=1 was already subtracted, so remove that from the slice
      outside <- h1/3 * 
                  ((-2 * d * sqrt(r1^2 - d^2)) +
                      ((d^3 / r1) * log(sqrt(r1^2 - d^2) + r1)) +
                      (r1^2 * acos(d/r1) - ((d^3 / r1) * log(d))))
        if (is.na(outside)){slicevol = slicevol}
        else {slicevol <- slicevol - outside}
      }
      allelevolume <- (conevol - 4*slicevol) + boxvol
    }
  return((allelevolume) / (25*25))
}


M = 0.1
S = 0.01
landscape_width=25
Fisher_wave_H <- function(M, # migration rate
                          S, # selection coefficient
                          landscape_width = 25) {
  landscape_area=landscape_width**2
  rmax=landscape_width/2 # radius @ which circle starts escaping the landscape
  Rmax=sqrt(2*(rmax**2)) # radius @ which circle encompasses landscape
  
  lambd=sqrt(M/S)
  V=M*sqrt(S) # allele velocity
  W=1/(4*sqrt(S)) # width of wavefront?????????????????????????????????????????????????
  hom_radii=seq(0, Rmax, V) # radius of the homozygous allele area circle after each generation
  het_radii=seq(0, Rmax+W, V) # radius of the heterozygous allele area circle after each generation
  hom_radii <- c(rep(0, length(het_radii) - length(hom_radii)), hom_radii) # add het only generations
  het_areas <- sapply(het_radii, allele_area)
  allele_areas <- het_areas
  het_freq <- mapply(allele_cone, het_radius=het_radii, hom_radius=hom_radii, hom_area=hom_areas)
  frequencies <- het_freq
  return(data.table(allele_areas, frequencies))
}

df <- Fisher_wave_H(M, S)
df$index <- seq(nrow(df))
ggplot(df) + geom_point(aes(x=index, y=frequencies))
```

```{r}
for (s in c(0.0001, 0.001, 0.01, 0.1)){
  df <- Fisher_wave_H(0.1, s)
  df$s <- s
  df$tick <- seq(nrow(df))
  assign(paste0('S_',s), df)
}

df <- rbind(`S_1e-04`, S_0.001, S_0.01, S_0.1)
ggplot(df) + theme_bw() + geom_line(aes(x=frequencies, y=allele_areas, group=s, color=s)) + scale_color_continuous(trans='log10')
```

```{r}
for (s in c(0.0001, 0.001, 0.01, 0.1)){
  df <- Fisher_wave_H(0.0001, s)
  df$s <- s
  df$tick <- seq(nrow(df))
  assign(paste0('S_',s), df)
}

df <- rbind(`S_1e-04`, S_0.001, S_0.01, S_0.1)
ggplot(df) + theme_bw() + geom_line(aes(x=frequencies, y=allele_areas, group=s, color=s)) + scale_color_continuous(trans='log10')
```



```{r}
for (s in c(0.0001, 0.001, 0.01, 0.1)){
  df <- Fisher_wave_H(0.1, s)
  df$s <- s
  df$tick <- seq(nrow(df))
  assign(paste0('S_',s), df)
}

df <- rbind(`S_1e-04`, S_0.001, S_0.01, S_0.1)
ggplot(df) + theme_bw() + geom_line(aes(x=frequencies, y=allele_areas, group=s, color=s)) + scale_color_continuous(trans='log10')
```